Differential Expression - HBCC - DLPFC - Sex

### 2020-08-19 individual-cohorts/split_by_analysis/HBCC_DLPFC_DE_Sex.Rmd
inital commit

---
title: "Differential Expression - MSSM-Penn-Pitt - DLPFC - Sex"
author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/MSSM-Penn-Pitt_DLPFC_DE.Rmd",
                                 parentId ="syn22043922",
                                 folderName = "Differential Expression - Sex")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)
library(R.utils)
library(ggplot2)
library(limma)
library(psych)
library(lme4)
library(biomaRt)
library(variancePartition)
library(edgeR)
library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')
library(WGCNA)

@@

```{r synapse.parameters, include=FALSE, cache=FALSE}
parentId = 'syn22043922';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Sex';

thisFileName <- 'MSSM-Penn-Pitt_DLPFC_DE.Rmd'

@@

# Download CQN counts
CQN_ID = "syn22045741"
ALL_USED_IDs = c(CQN_ID)
CQN.GENE_EXPRESSION = downloadFile(CQN_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id") %>% 
  data.matrix()

# Download filtered counts
COUNTS_ID = "syn22045729"
ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
NEW.COUNTS = downloadFile(COUNTS_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id")

# Get covariates
COVAR_ID = "syn22045728"
ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
COVARIATES = downloadFile(COVAR_ID)

# dropped individual id 
MD_ID = "syn16816488"
ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
METADATA = downloadFile(MD_ID) %>% 
  dplyr::select(Individual_ID, Sample_RNA_ID)

COVARIATES = right_join(METADATA, COVARIATES, by = c("Sample_RNA_ID" = "SampleID")) %>% 
  tibble::column_to_rownames(var = "Sample_RNA_ID")

# Set variable class
FactorCovariates <- c("Individual_ID", "Institution", "LibraryBatch", "Dx", "Reported_Gender", "RibozeroBatch", "FlowcellBatch")
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)

@@

# Set covariates to adjust for from model selection
adjust.covars <- c("Sex", "Dx", "ageOfDeath", "RIN2", "RIN", "IntronicRate", "IntragenicRate")
random_effect <- c("Institution")
```

@@

```{r iterative.normalisation, results='asis'}
writeLines(glue::glue('Using following covariates in the model:',
                 glue::glue_collapse(adjust.covars, sep = ", ", last = " and "),
                 ' as fixed effects and',
                 glue::glue_collapse(random_effect, sep = ", ", last = " and "),' chosen as random effects'))

# Post adjusted formula
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + "), " + ", glue::glue_collapse(glue::glue("(1|{random_effect})"), sep = " + "))

@@

col_type <- dplyr::select(COVARIATES, c(adjust.covars, random_effect)) %>%
  dplyr::summarise_all(class) %>% 
  tidyr::gather(variable, class)

# Model categorical variables as random to report variance fractions
formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
  switch(col_type$class[i], 
    "factor" = paste0('(1|', col_type$variable[i], ')'),
    "numeric" = col_type$variable[i])
  })

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Institution, shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

```{r decompose.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Sex','Institution')])

@@

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Sex','Institution')]))
```

@@

### Differential expression analysis (Female - Male)
Genes that are differentially expressed at an FDR <= 0.05 are:
```{r set-formula}
adjust.covars <- adjust.covars[!(adjust.covars %in% "RIN2")]
formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + "), " + ", glue::glue_collapse(glue::glue("(1|{random_effect})"), sep = " + "))    
```

@@

```{r diffExp, fig.height=10, fig.width=15}
L = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("SexFemale", "SexMale"))
contrast <- c(glue::glue_collapse(names(L[L != 0]), "-"))

@@

# Store adjusted fit linear mixed model object as RDS - Sex
saveRDS(ADJUSTED.FIT, file = 'CMC_DLPFC_Sex_dream-output.rds')
ADJ_OBJ = File('CMC_DLPFC_Sex_dream-output.rds',
               name = 'Dream Output object - Sex', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)
```

>>>>

 ---
title: "Differential Expression - HBCC - DLPFC - Sex"
author: "Kelsey Montgomery, Gabriel Hoffman, Thanneer Perumal"
date: "`r date()`"
output: html_document
---

@@

knit2synapse::createAndKnitToFolderEntity(file = "individual-cohorts/HBCC_DLPFC_DE.Rmd",
                                 parentId ="syn22050019",
                                 folderName = "Differential Expression - Sex")
```

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## Load required libraries
library(CovariateAnalysis) # get the package from devtools::install_github('th1vairam/CovariateAnalysis@dev')
library(data.table)
library(tidyr)
library(plyr)
library(dplyr)
library(stringr)
library(R.utils)
library(ggplot2)
library(limma)
library(psych)
library(lme4)
library(biomaRt)
library(variancePartition)
library(edgeR)
library(synapser)
library(knitr)
library(githubr) # get the package from devtools::install_github('brian-bot/githubr')

@@

```{r synapse.parameters, include=FALSE, cache=FALSE}
parentId = 'syn22050019';
activityName = 'Differential Expression';
activityDescription = 'Differential Expression - Sex';

thisFileName <- 'HBCC_DLPFC_DE.Rmd'

@@

# Download CQN counts
CQN_ID = "syn22082460"
ALL_USED_IDs = c(CQN_ID)
CQN.GENE_EXPRESSION = downloadFile(CQN_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id") %>% 
  data.matrix()

# Download filtered counts
COUNTS_ID = "syn22082459"
ALL_USED_IDs = c(ALL_USED_IDs, COUNTS_ID)
NEW.COUNTS = downloadFile(COUNTS_ID) %>% 
  tibble::column_to_rownames(var = "ensembl_gene_id")

# Get covariates
COVAR_ID = "syn22082458"
ALL_USED_IDs = c(ALL_USED_IDs, COVAR_ID)
COVARIATES = downloadFile(COVAR_ID) %>% 
  tibble::column_to_rownames(var = "SampleID")

# Set variable class
FactorCovariates <- c("Individual_ID", "Reported_Gender", "LibraryBatch", "Dx", "RibozeroBatch", "FlowcellBatch")
COVARIATES[,FactorCovariates] = lapply(COVARIATES[,FactorCovariates], factor)

@@

# Set covariates to adjust for from model selection
adjust.covars <- c("Sex", "Dx", "ageOfDeath", "RIN", "IntronicRate", "IntragenicRate", "IntergenicRate")
```

@@

```{r iterative.normalisation, results='asis'}
writeLines(glue::glue('Using following covariates in the model:',
                 glue::glue_collapse(adjust.covars, sep = ", ", last = " and "),
                 ' as fixed effects.'))

col_type <- dplyr::select(COVARIATES, c(adjust.covars)) %>%
  dplyr::summarise_all(class) %>% 
  tidyr::gather(variable, class)

# Scale continuous variables
adjust.covars = sapply(1:length(col_type$class), function(i){
  switch(col_type$class[i], 
    "factor" = paste0(col_type$variable[i]),
    "numeric" = paste0('scale(', col_type$variable[i], ')')
  )
  })

# Post adjusted formula
formula <- glue::glue("~ ", glue::glue_collapse(adjust.covars, sep = " + "))

@@

# Model categorical variables as random to report variance fractions
formula_for_variance_fractions = sapply(1:length(col_type$class), function(i){
  switch(col_type$class[i], 
    "factor" = paste0('(1|', col_type$variable[i], ')'),
    "numeric" = paste0('scale(', col_type$variable[i], ')')
  )
  })

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(shape=Tissue, size=ageOfDeath))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

```{r decompose.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Sex'), drop = FALSE])

@@

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Sex'), drop = FALSE]))
```

@@

### Differential expression analysis (SCZ - Control - Other)
Genes that are differentially expressed at an FDR <= 0.05 are:
```{r set-formula}
formula <- glue::glue("~ 0 + ", glue::glue_collapse(adjust.covars, sep = " + "))    
```

@@

```{r diffExp, fig.height=10, fig.width=15}
L = variancePartition::getContrast(exprObj = VOOM.GENE_EXPRESSION, 
                                   formula = formula, 
                                   data = COVARIATES, 
                                   coefficient = c("SexFemale", "SexMale"))

contrast <- c(glue::glue_collapse(names(L[L != 0]), "-"))

@@

# Store adjusted fit linear mixed model object as RDS - Sex
saveRDS(ADJUSTED.FIT, file = 'CMC_DLPFC_Sex_dream-output.rds')
ADJ_OBJ = File('CMC_DLPFC_Sex_dream-output.rds',
               name = 'Dream Output object - Sex', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)

# Store variancePartition
saveRDS(sortCols(violin_plot), file = 'CMC_DLPFC_Sex_variancePartition.rds')
ADJ_OBJ = File('CMC_DLPFC_Sex_variancePartition.rds',
               name = 'Variance Partition result', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)

# Store voom object
saveRDS(VOOM.GENE_EXPRESSION, file = 'CMC_DLPFC_Sex_voom.rds')
ADJ_OBJ = File('CMC_DLPFC_Sex_voom.rds',
               name = 'voom object', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)

# Store contrasts
saveRDS(L, file = 'CMC_DLPFC_Sex_contrasts.rds')
ADJ_OBJ = File('CMC_DLPFC_Sex_contrasts.rds',
               name = 'contrast matrix', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)

# Store contrasts
saveRDS(formula, file = 'CMC_DLPFC_Sex_formula.rds')
ADJ_OBJ = File('CMC_DLPFC_Sex_formula.rds',
               name = 'formula', 
               parentId = CODE$properties$id)
ADJ_OBJ = synStore(ADJ_OBJ, used = ALL_USED_IDs, activityName = activityName,
                   executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(ADJ_OBJ, annotations = all.annotations)

```
